---
title: "Immediate Effects of Storms on Population Health and Economics, 1950 - 2011"
output: html_document
---

## Synopsis

Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

```{r, message = FALSE}
library(dplyr)
```

## Data Processing

There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

First the data are loaded in from the bz2 compressed csv provided. This operation is large and should be cached.

```{r, cache = T}
stormdata <- read.csv("StormData.csv.bz2")
```

#### Population Health

To see how specific event types affect population health, fatalities, injuries, and their sum should be aggregated against event types:

```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, data = stormdata, FUN = "sum")
injuries <- aggregate(INJURIES ~ EVTYPE, data = stormdata, FUN = "sum")
combined <- aggregate(INJURIES + FATALITIES ~ EVTYPE, data = stormdata, FUN = "sum")
combined <- cbind(combined, fatalities$FATALITIES, injuries$INJURIES)

colnames(combined) <- c('EVTYPE','SUM','FATALITIES','INJURIES')
combined <- arrange(combined, -SUM)
```

We can now easily see the storm event types with the largest affects on population help:

```{r}
head(combined, 10)
```

#### Economic Impact

This data is much more difficult to work with because ~~~

```{r}
econdata <- stormdata[c('EVTYPE','CROPDMG','CROPDMGEXP','PROPDMG','PROPDMGEXP')]
replace <- function(x) {
  sub("\\?",0,x)
  sub("\\+",0,x)
  sub("-",0,x)
}


```

## Results



Present results
At least one plot, no more than 3
