---
title: "Immediate Effects of Storms on Population Health and Economics, 1950 - 2011"
output: github_document
---

## Synopsis

Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
```

## Data Processing

There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

First the data are loaded in from the bz2 compressed csv provided. This operation is large and should be cached.

```{r, cache = T}
stormdata <- read.csv("StormData.csv.bz2")
```

#### Population Health

To see how specific event types affect population health, fatalities, injuries, and their sum should be aggregated against event types:

```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, data = stormdata, FUN = "sum")
injuries <- aggregate(INJURIES ~ EVTYPE, data = stormdata, FUN = "sum")
combined <- aggregate(INJURIES + FATALITIES ~ EVTYPE, data = stormdata, FUN = "sum")
combined <- cbind(combined, fatalities$FATALITIES, injuries$INJURIES)

colnames(combined) <- c('EVTYPE','SUM','FATALITIES','INJURIES')
combined <- arrange(combined, -SUM)
```

We can now easily see the storm event types with the largest affects on population help:

```{r}
head(combined, 10)
```

#### Economic Impact

This data is much more difficult to work with because there is no single variable describing property damage or crop damage, but rather each of those variables has a base value and a multiplier. CROPDMGEXP and PROPDMGEXP modify the value of CROPDMG and PROPDMG respectively. Furthermore, the multipliers are encoded in a strange way. When can see the modifiers with the following code:

```{r}
econdata <- stormdata[c('EVTYPE','CROPDMG','CROPDMGEXP','PROPDMG','PROPDMGEXP')]
econdata$CROPDMGEXP <- toupper(econdata$CROPDMGEXP)
econdata$PROPDMGEXP <- toupper(econdata$PROPDMGEXP)

sort(unique(econdata$PROPDMGEXP))

```

We can replace these modifiers with their corresponding numeric values and remove incomplete cases. I've queried the first 5 rows of the multipliers to confirm the substitution worked successfully. Note that this code block is computationally expensive and benefits from caching.

```{r, cache = T}
EXPkey <-  c("\"\"" = 10^0,
             "-" = 10^0, 
             "+" = 10^0,
             "0" = 10^0,
             "1" = 10^1,
             "2" = 10^2,
             "3" = 10^3,
             "4" = 10^4,
             "5" = 10^5,
             "6" = 10^6,
             "7" = 10^7,
             "8" = 10^8,
             "9" = 10^9,
             "H" = 10^2,
             "K" = 10^3,
             "M" = 10^6,
             "B" = 10^9)
econdata$CROPDMGEXP <- EXPkey[econdata$CROPDMGEXP]
econdata <- subset(econdata, !is.na(econdata$CROPDMGEXP))

econdata$PROPDMGEXP <- EXPkey[econdata$PROPDMGEXP]
econdata <- subset(econdata, !is.na(econdata$PROPDMGEXP))

head(econdata$CROPDMGEXP,5)
head(econdata$PROPDMGEXP,5)
```

We can now multiply each of the numeric values by their multiplier in order to obtain their true values.

```{r}
econdata$crop.damage <- econdata$CROPDMG * econdata$CROPDMGEXP
econdata$property.damage <- econdata$PROPDMG * econdata$PROPDMGEXP

crop.sum <- aggregate(crop.damage ~ EVTYPE, data = econdata, FUN = "sum")
property.sum <- aggregate(property.damage ~ EVTYPE, data = econdata, FUN = "sum")
econ.sums <- cbind(crop.sum, property.sum$property.damage)

colnames(econ.sums) <- c("event.type","crop.damage","property.damage")
econ.sums$total.damage <- econ.sums$crop.damage + econ.sums$property.damage
econ.sums <- arrange(econ.sums, -total.damage)

head(econ.sums, 5)
```

## Results

#### Population Health
```{r}
x <- combined[1:7,]
x$EVTYPE <- factor(x$EVTYPE, levels=x$EVTYPE) # So ggplot doesnt alphabatize x-axis 
ggplot(x, aes(EVTYPE, SUM)) + 
  theme_bw() +
  geom_bar(stat="identity") + 
  xlab(NULL) +
  ylab("Injuries + Fatalities") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

Tornadoes are responsible for the most personal harm by a wide margin. It's interesting to note that floods and flash floods are classified seperately. We can also view the total instances of personal harm from all sources:

```{r}
sum(combined$SUM)
```

#### Economic Impact

```{r}
x <- econ.sums[1:7,]
x$event.type <- factor(x$event.type, levels=x$event.type) # So ggplot doesnt alphabatize x-axis 
ggplot(x, aes(event.type, total.damage)) + 
  theme_bw() +
  geom_bar(stat="identity") + 
  xlab(NULL) +
  ylab("Property Damage + Crop Damage") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

Flooding is responsible for the most economic damage by a wide margin. It's interesting to note that hurricane/typhoon and typhoon are clasified seperately. We can also view the total economic damage from all sources:

```{r}
sum(econ.sums$total.damage)
```
