---
title: "Analysis of Activity Monitoring Data"
author: "Sawyer Welden"
date: "July 28, 2017"
output: html_document
---

## Load in the data and required libraries

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(lubridate)
```

This analysis requires the ggplot2 and lubridate packages

```{r}
data <- read.csv("activity.csv")
```

## Total Steps per Day

Aggregate and plot total steps per day
```{r}
steps.by.date <- aggregate(steps ~ date, data = data, FUN = sum)
ggplot(data = steps.by.date, aes(steps)) + 
  geom_histogram(bins = 30) + 
  xlab("Steps") + 
  ylab("Count")
```

Mean and Median steps per day
```{r}
mean(steps.by.date$steps)
median(steps.by.date$steps)
```

## Analysing activity intervals

Average steps per 5-minute interval
```{r}
steps.by.time <- aggregate(steps ~ interval, data = data, FUN = mean)
ggplot(data = steps.by.time, aes(interval, steps)) + 
  geom_line() + 
  xlab("Interval") + 
  ylab("Steps")
```

Highest Activity Interval
```{r}
steps.by.time[which.max(steps.by.time$steps),]$interval
```
The interval with the highest activity was 8:35 - 8:40

## Imputing Missing Data

First copy the original dataset
```{r}
data.imputed <- data
```

Then change all NA's to the mean value (excluding NA's)
```{r}
mean.steps <- mean(data$steps, na.rm = TRUE)
data.imputed[is.na(data.imputed)] <- mean.steps
```

Histogram of steps per day including imputed values
```{r}
imputed.by.date <- aggregate(steps ~ date, data = data.imputed, FUN = sum)
ggplot(data = imputed.by.date, aes(steps)) + 
  geom_histogram(bins = 30) + 
  xlab("Steps") + 
  ylab("Count")
```

## Weekdays vs. Weekends
Determine which days are weekends and include that in aggregation
```{r}
isWeekday <- function(x) {
  day <- wday(x)
  return(day %in% seq(1,5))
}
data$is.weekday <- isWeekday(data$date)
agr <- aggregate(steps ~ interval + is.weekday, data = data, FUN = mean)
agr$is.weekday <- as.factor(agr$is.weekday)
levels(agr$is.weekday) <- c('Weekday','Weekend')
```
Plot activity 

```{r}
ggplot(data = agr, aes(interval, steps)) + 
  geom_line() + 
  xlab("Interval") + 
  ylab("Steps") + 
  facet_grid(is.weekday ~ .)
```
